---
title: "Diet and Sleep vs. Condition"
author: "Waveley Qiu"
date: "`r format(Sys.time(), '%Y-%m-%d')`"
output: word_document
header-includes:
  - \usepackage{mathrsfs}
  - \usepackage{amsfonts}  
---

```{r setup, include=FALSE}
set.seed(20220417)
library(tidyverse)
library(lme4)
library(lmerTest)
library(kableExtra)
# set knitr defaults
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.align = "center",
  fig.width = 6,
  fig.asp   = .6,
  out.width = "90%",
  cache = TRUE
)

# set theme defaults
theme_set(
  theme_bw() +
    theme(
      legend.position = "bottom",
      plot.title    = element_text(hjust = 0.5),
      plot.subtitle = element_text(hjust = 0.5),
      plot.caption  = element_text(hjust = 0.0)
    )
)

# set color scale defaults
options(
  ggplot2.continuous.colour = "viridis",
  ggplot2.continuous.fill   = "viridis"
)
scale_colour_discrete = scale_colour_viridis_d
scale_fill_discrete   = scale_fill_viridis_d
```

## Background

This analysis seeks to understand the influence that alignment condition has on two categories of lifestyle outcomes: dietary intake and sleep measures.

## Data

There are 8 subjects in this study. All subjects have complete records for all outcome measures of interest. 

In assessing the influence that alignment condition has on dietary intake, each subject was measured at two timepoints: week 3 and week 6. As each subject underwent both alignment conditions (aligned and misaligned), each subject had four total records in that round of data collection.

In assessing the influence that alignment condition has on dietary intake, each subject was measured at six timepoints: weeks 1, 2, 3, 4, 5, and 6. As each subject underwent both alignment conditions (aligned and misaligned), each subject had twelve total records in that round of data collection.

## Variables

The predictors in both arms of this analysis (diet outcomes and sleep outcomes) are the same -- we are interested in one's alignment condition (aligned/misaligned), the week of measurement, as well as the interaction of condition and week. In this analysis, week is treated as a continuous variable.

In investigating the influence that one's alignment condition may have on their sleep, we are interested in the following outcome variables: food weight, energy, food weight (no beverages), energy (no beverages), energy density, protein, fat, carbohydrate, plant protein, animal protein, sugar, fiber, calcium, magnesium, potassium, sodium, zinc, vitamin C, vitamin B6, folate, vitamin B12, saturated fat, monounsaturated fat, polyunsaturated fat, unsaturated fat, vitamin D, fruits, vegetables, fruits and vegetables, eggs, nuts, dairy, and added sugar. 

In investigating the influence that one's alignment condition may have on their diet, we are interested in the following predictors: latency, efficiency, TIB, TST, WASO, number of awakenings, awakening length, movement index, SFI, and fragmentation index.

## Methodology

For both portions of this analysis, linear mixed effect models, with random intercepts for each subject, will be constructed to regress each outcome variable in turn on the following predictors: condition, week, the interaction of condition and week. 

In all models constructed, phase (coded as a factor variable, taking in values 1 and 2) will be included in the model if it is deemed to be significant. In addition, the diet portion of this study, energy (recorded as the variable "en" in the dataset) will be added as a predictor in all models, excluding the model in which energy itself is the outcome. 

\newpage

## Results

### Diet and Condition

The results of regressing diet outcome variables on condition, week, the interaction of condition and week, energy (if applicable), and phase (if determined to be a significant predictor) are as follows:

```{r, echo = FALSE}
diet <- readxl::read_excel("data/CM Study Diet and Sleep_10.5.2022_HB.xlsx", sheet = "Diet") %>%
  janitor::clean_names() %>%
  mutate(
    sex = as.factor(sex),
    phase = as.factor(phase)
  )

outs1 <-
  diet %>%
  select(
    food_wt_ttl, en, food_wt_f, en_food, 
    ed, prot, fat, cho, prot_plant,
    prot_ani, sugar, fiber, ca, mg, k, 
    na, zn, vit_c, vit_b6, folate,
    vit_b12, sfa, mufa, pufa, ufa, vit_d, 
    fruit, veg, f_v, egg, nuts, dairy, 
    added_sugar
  )

preds1 <-
  diet %>%
  select(
    condition, week, phase, en, id
  )

loop_diet_models <- function(outcomes = outs1, preds = preds1){
      
    condition <- preds$condition
    week <- preds$week
    phase <- preds$phase
    en <- preds$en
    id <- preds$id
    
    outcome_tib <- 
      tibble()
    
      for (j in 1:ncol(outcomes)) {
        
      outcome_name <- outcomes[,j] %>% names()
        
        #print(paste0(outcome_name))
        
        cur_outcome <- outcomes %>% pull(outcome_name)

        if ("en" %in% outcome_name) {
          cur_formula <- "cur_outcome ~ condition + week + condition*week + phase + (1 | id)"
        }
        else if (!("en" %in% outcome_name)) {
          cur_formula <- "cur_outcome ~ condition + week + condition*week + en + phase + (1 | id)"
        }

        first_mem <- lmer(cur_formula)
        saved_sum <- first_mem %>% summary()
        p_vals <- saved_sum$coefficients[,5] # p-value column
        save_covs <- p_vals[p_vals < 0.05] %>% names() # want to know if phase is significant

        fin_formula <- cur_formula
        
        if (!("phase2" %in% save_covs) & "en" %in% outcome_name) {
          fin_formula <- "cur_outcome ~ condition + week + condition*week + (1 | id)"
        }
        else if (!("phase2" %in% save_covs) & !("en" %in% outcome_name)) {
          fin_formula <- "cur_outcome ~ condition + week + condition*week + en + (1 | id)"
        }
        
        #print(fin_formula)
        final_mem <- lmer(fin_formula)
        mod_sum <- summary(final_mem)
        
        # save model summary output in list
        mod_sum_output <- capture.output(summary(final_mem))

        # extract relevant portions of summary output
        first_ind <- mod_sum_output %>% match("Fixed effects:")
        last_ind <- mod_sum_output %>% match("Correlation of Fixed Effects:")
        
        mod_sum_output_trunc <-
          mod_sum_output[(which(first_ind == 1) + 1):(which(last_ind == 1) - 2)]
        
        mod_sum_coef <- mod_sum$coefficients
        mod_estimates <- mod_sum_coef[,1]

        cur_tib <-
          tibble(
            outcome = outcome_name,
            model_sum = list(mod_sum_output_trunc)
            
          )
        
        outcome_tib <- outcome_tib %>% bind_rows(cur_tib)
    }
  return(outcome_tib)
}

write_mod <- function(out, 
                      mod_sum){
  
  new_vals <- 
    mod_sum %>% 
    unlist() %>%
    round(digits = 4) 
  
  rhs <- tibble(
    name = new_names,
    vals = new_vals
  ) %>%
    mutate(
      sign = sign(vals),
      new_vals = abs(vals),
      new_term = ifelse(name == "", 
                        ifelse(sign == 1, paste(new_vals), paste("-", new_vals)), 
                       ifelse(sign == 1, 
                              paste0(" + ", new_vals, "(", name, ")"), 
                              paste0(" - ", new_vals, "(", name, ")")))
    ) %>% pull(new_term) %>% paste(collapse = "")
  
  lhs <- out %>% toupper()
  fin_mod <- paste0(lhs, " = ", rhs)
return(fin_mod)
}

diet_models <- loop_diet_models(outs1, preds1)

for (i in 1:nrow(diet_models)) {
  cur_df <- diet_models[i,]
  writeLines(paste0("Model ", i, ": ", cur_df$outcome, " vs. condition"))
  writeLines(unlist(lapply(cur_df$model_sum, paste)))
  writeLines("\n")
}
```

\newpage

## Sleep and Condition

The results of regressing sleep outcome variables on condition, week, the interaction of condition and week, and phase (if determined to be a significant predictor) are as follows:

```{r, echo = FALSE}
sleep <- readxl::read_excel("data/CM Study Diet and Sleep_10.5.2022_HB.xlsx", sheet = "Sleep") %>%
  janitor::clean_names() %>%
  mutate(
    sex = as.factor(sex),
    phase = as.factor(phase)
  )

outs2 <-
  sleep %>%
  select(
    latency, efficiency, tib, tst, waso, 
    number_of_awakenings, awakening_length, 
    movement_index, sfi, fragmentation_index
  )

preds2 <-
  sleep %>%
  select(
    condition, week, phase, id
  )

loop_sleep_models <- function(outcomes = outs2, preds = preds2){
      
    condition <- preds$condition
    week <- preds$week
    phase <- preds$phase
    id <- preds$id
    
    outcome_tib <- 
      tibble()
    
      for (j in 1:ncol(outcomes)) {
        
        outcome_name <- outcomes[,j] %>% names()
        
        #print(paste0(outcome_name))
        
        cur_outcome <- outcomes %>% pull(outcome_name)
        cur_formula <- "cur_outcome ~ condition + week + condition*week + phase + (1 | id)"
        
        #print(cur_formula)
        
        first_mem <- lmer(cur_formula)
        saved_sum <- first_mem %>% summary()
        p_vals <- saved_sum$coefficients[,5] # p-value column
        save_covs <- p_vals[p_vals < 0.05] %>% names() # want to know if phase is significant
        
        fin_formula <- cur_formula

        if (!("phase2" %in% save_covs)) {
          fin_formula <- "cur_outcome ~ condition + week + condition*week + (1 | id)"
        }
        
        #print(fin_formula)
        final_mem <- lmer(fin_formula)
        mod_sum <- summary(final_mem)
        
        # save model summary output in list
        mod_sum_output <- capture.output(summary(final_mem))

        # extract relevant portions of summary output
        first_ind <- mod_sum_output %>% match("Fixed effects:")
        last_ind <- mod_sum_output %>% match("Correlation of Fixed Effects:")
        
        mod_sum_output_trunc <-
          mod_sum_output[(which(first_ind == 1) + 1):(which(last_ind == 1) - 2)]
        
        mod_sum_coef <- mod_sum$coefficients
        mod_estimates <- mod_sum_coef[,1]

        cur_tib <-
          tibble(
            outcome = outcome_name,
            model_sum = list(mod_sum_output_trunc)
            
          )
        
        outcome_tib <- outcome_tib %>% bind_rows(cur_tib)
    }
  return(outcome_tib)
}

sleep_models <- loop_sleep_models(outs2, preds2)

for (i in 1:nrow(sleep_models)) {
  cur_df <- sleep_models[i,]
  writeLines(paste0("Model ", i, ": ", cur_df$outcome, " vs. condition"))
  writeLines(unlist(lapply(cur_df$model_sum, paste)))
  writeLines("\n")
}
```

\newpage

## GitHub Repository

All code for this report can be found in [this Github repository](https://github.com/waveley/diet_sleep).


