---
title: "Diet and Sleep (HEI)"
author: "Waveley Qiu"
date: "`r format(Sys.time(), '%Y-%m-%d')`"
output: pdf_document
header-includes:
  - \usepackage{mathrsfs}
  - \usepackage{amsfonts}  
---

```{r setup, include=FALSE}
set.seed(20220119)
library(tidyverse)
library(lme4)
library(lmerTest)
library(kableExtra)
# set knitr defaults
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.align = "center",
  fig.width = 6,
  fig.asp   = .6,
  out.width = "90%",
  cache = TRUE
)

# set theme defaults
theme_set(
  theme_bw() +
    theme(
      legend.position = "bottom",
      plot.title    = element_text(hjust = 0.5),
      plot.subtitle = element_text(hjust = 0.5),
      plot.caption  = element_text(hjust = 0.0)
    )
)

# set color scale defaults
options(
  ggplot2.continuous.colour = "viridis",
  ggplot2.continuous.fill   = "viridis"
)
scale_colour_discrete = scale_colour_viridis_d
scale_fill_discrete   = scale_fill_viridis_d
```

## Background

This analysis seeks to assess whether diet influences sleep.

## Data

Data from CM and DQ were combined for the purpose of this analysis. In order to establish a temporal order, each data about each patients' diets was collected prior to data on each patients' sleep qualities. 

## Variables

We are interested in the following predictors: caffeine, whole grains, refined grains, poultry, soy, legumes, red and processed meat, seaffod, saturated/unsaturated fat ratio, long chain omega 3 fatty acids, and the healthy eating index (HEI) total score. We are interested in the following outcome variables that measure sleep: sleep efficiency, total sleep time, wake after sleep onset, and sleep fragmentation index. 

Some of the variables required by the provided `hei2015.score.macro` SAS macro needed to be derived. The variables and their derivations are as follows:

- VTOTALLEG (total veg plus legumes in cup eq.): V_TOTAL + V_LEGUMES

- VDRKGRLEG (intake of dark green veg plus legumes in cup eq.): V_DRKGR + V_LEGUMES

- FWHOLEFRT (intake of whole fruit in cup eq.) = F_TOTAL - F_JUICE

- PFALLPROTLEG (intake of total protein [including legumes] in oz. eq.): PF_TOTAL + PF_LEGUMES

- PFSEAPLANTLEG (intake of seafood, fish, and protein [includes legumes] in oz. eq.): SEAFOOD + PF_SOY + PF_NUTSDS + PF_LEGUMES

- MONOPOLY (grams of mono fat plus poly fat): MFAT + PFAT

In addition, energy (kcal), age, sex, and BMI will be used as covariates. Condition will not be used as a covariate in the models, as all subjects in the provided data were under the same condition ("Aligned").

## Methodology

We will construct linear mixed effect models to conduct a preliminary assessment of the relationships between all variables of interest. First, we will regress the outcome variable of interest on one predictor and the four covariates, adding a random intercept for subject ID. We will then evaluate the significance of the coefficients associated with each of the three covariates. If any of the coefficients associated with age, sex, BMI, or condition are determined to not be significant (i.e., p > 0.05), the associated covariate will be removed from the model. The truncated model will then be run again and its result saved.

\newpage

## Results

### Diet and Sleep

The following table presents the results of regressing sleep outcome variables on diet predictors:

```{r, echo = FALSE}
diet_sleep <- 
  readxl::read_excel("data/CM + DQ Study Diet and Sleep_Added Diet Variables and HEI_12.14.2022_with_hei.xlsx", sheet = "Diet-sleep HEI") %>%
  janitor::clean_names() %>%
  mutate(
    sex = as.factor(sex)
  )

outs1 <-
  diet_sleep %>%
  select(
    se,
    tst,
    waso, 
    sfi
  )
covs1 <-
  diet_sleep %>%
  select(
    sex, age, bmi, id, kcal
  )
preds1 <- 
  diet_sleep %>%
  select(
    caff, g_whole, g_refined, pf_poult,
    pf_soy, pf_legumes, red_processed_meat, 
    seafood, saturated_unsaturated_ratio, 
    omega3_fa, hei2015_total_score
  )

loop_models <- function(outcomes = outs1, predictors = preds1, covariates = covs1){
      
    outcome_tib <- 
      tibble()
    
    age <- covariates$age
    sex <- covariates$sex
    bmi <- covariates$bmi
    condition <- covariates$condition
    id <- covariates$id
    kcal <- covariates$kcal
    
    for (i in 1:ncol(predictors)) {
      for (j in 1:ncol(outcomes)) {
        
        pred_name <- predictors[,i] %>% names()
        outcome_name <- outcomes[,j] %>% names()
        
       # print(paste0(outcome_name, " vs ", pred_name))
        
        cur_outcome <- outcomes %>% pull(outcome_name)
        cur_predictor <- predictors %>% pull(pred_name)
        
        cur_cov_vec <- c("age", "sex", "bmi") # all covariates 
        
        cur_formula <-
          paste0("cur_outcome ~ ", 
                 paste0(cur_cov_vec, sep = " + ", collapse = ""),
                 "kcal + cur_predictor + (1 | id)"
                 )
        
        first_mem <- lmer(cur_formula)
        saved_sum <- first_mem %>% summary()
        p_vals <- saved_sum$coefficients[,5] # p-value column
        save_covs <- p_vals[p_vals < 0.05] %>% names()
        save_covs <- save_covs %>% str_replace_all("sex2", "sex") # only keep covariates with p-value < 0.05
        updated_cov_vec <- save_covs[!(save_covs %in% c("(Intercept)"))] # dont want intercept
        
        fin_formula <-
          paste0("cur_outcome ~ ", 
                 paste0(updated_cov_vec, sep = " + ", collapse = ""),
                 "kcal + cur_predictor + (1 | id)"
                 )
        final_mem <- lmer(fin_formula)
       # tidy_mem <- final_mem %>% texreg()
        mod_sum <- summary(final_mem)
        
        # save model summary output in list
        mod_sum_output <- capture.output(summary(final_mem))
        
        # extract relevant portions of summary output
        first_ind <- mod_sum_output %>% match("Fixed effects:")
        last_ind <- mod_sum_output %>% match("Correlation of Fixed Effects:")
        
        mod_sum_output_trunc <-
          mod_sum_output[(which(first_ind == 1) + 1):(which(last_ind == 1) - 2)]
        
        # replace "cur_predictor" with current predictor name
         
         
        mod_sum_pred_row <- mod_sum_output_trunc %>% str_locate("cur_predictor")
        mod_sum_output_trunc[which(mod_sum_pred_row[,1] == 1)] <- 
          str_replace(mod_sum_output_trunc[which(mod_sum_pred_row[,1] == 1)],
                      "cur_predictor",
                      str_trunc(paste0(pred_name, "             "), 13, side = "right", ellipsis = ""))
        
        mod_sum_coef <- mod_sum$coefficients
        mod_estimates <- mod_sum_coef[,1]
        pred_p_val <- mod_sum_coef["cur_predictor", 5]
        
        cur_tib <-
          tibble(
            predictor = pred_name,
            outcome = outcome_name,
            model_sum = list(mod_sum_output_trunc),
            model_summary = list(mod_sum_coef),
            model_estimates = list(mod_estimates),
            predictor_pval = pred_p_val %>% round(digits = 3),
            sig = if_else(predictor_pval < 0.05, "***", "")
            
          )
        
        outcome_tib <- outcome_tib %>% bind_rows(cur_tib)
    }
  }
  return(outcome_tib)
}

write_mod <- function(pred = test$predictor, 
                      out = test$outcome, 
                      mod_sum = test$model_estimates){
  
  new_names <- 
    mod_sum %>% 
    unlist() %>%
    names() %>% 
    str_replace_all("cur_predictor", pred) %>%
    str_replace_all('\\(Intercept\\)',"") %>% toupper()
  
  new_vals <- 
    mod_sum %>% 
    unlist() %>%
    round(digits = 4) 
  
  rhs <- tibble(
    name = new_names,
    vals = new_vals
  ) %>%
    mutate(
      sign = sign(vals),
      new_vals = abs(vals),
      new_term = ifelse(name == "", 
                        ifelse(sign == 1, paste(new_vals), paste("-", new_vals)), 
                       ifelse(sign == 1, 
                              paste0(" + ", new_vals, "(", name, ")"), 
                              paste0(" - ", new_vals, "(", name, ")")))
    ) %>% pull(new_term) %>% paste(collapse = "")
  
  lhs <- out %>% toupper()
  fin_mod <- paste0(lhs, " = ", rhs)
return(fin_mod)
}

do_write_mods <- function(df) {
  
 fin_df <- df %>% 
  mutate(
    fin_mods = pmap(list(predictor, outcome, model_sum), new_write_mods)
  ) %>%
  select(predictor, outcome, fin_mods, predictor_pval, sig) %>%
  unnest(fin_mods)
 
 return(fin_df)
}

diet_sleep_models <- loop_models(outs1, preds1, covs1)

for (i in 1:nrow(diet_sleep_models)) {
  cur_df <- diet_sleep_models[i,]
  writeLines(paste0("Model ", i, ": ", cur_df$outcome, " vs. ", cur_df$predictor))
  writeLines(unlist(lapply(cur_df$model_sum, paste)))
  writeLines("\n")
}
```

\newpage

## GitHub Repository

All code for this report can be found in [this Github repository](https://github.com/waveley/diet_sleep).


